### 2017-12-02 thesis/contents/reviews/mri.tex
checkpoint on edit

%\graphicspath{{D:/project/thesis_preparation/PhDtemplateLATEX_V2/ch_Background_Work/figures/}}
\section{Magnetic Resonance Imaging}
Magnetic Resonance Imaging (MRI) is a medical imaging technology that allows non-invasive visualization of the body using magnetic fields. It is a key enabler of modern healthcare and medical research, and is used in the detection and diagnosis of neuropathology.

MRI is based on the principles of nuclear magnetic resonance (NMR), where atomic nuclei, when placed in a magnetic field, can absorb and emit electromagnetic radiation. A MRI machine applies a primary magnetic field (B0) to magnetize and align the hydrogen nuclei in the body. A radio frequency (RF) field is subsequently introduced perpendicular to B0, and reorients the spin axis. This causes the nuclei to produce NMR signals detectable by the MRI scanner. Based on the precession frequency decay, the signals are recorded to construct an image of the scanned area based on the contrasts in water and fat concentrations.

@@

\begin{figure}[h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=3.5in]{mri/spin.pdf}
\end{tabular}
\caption{Spin of a nucleus has a magnetic moment vector $\mu$, which behaves similar to a magnet with north (N) and south (S) poles.}
\label{fig:spin}
\end{center}
\end{figure}

@@

As protons aligned in the parallel orientation are more stable, the number of spins under this state is more than that under anti-parallel orientation. This number of difference is called spin excess and the magnetic moments of the excess spins add to form the net magnetization vector $M_z$. When the spin system is at equilibrium, net magnetization lies along the z axis, parallel to $B_0$.

@@

\subsubsection{Relaxation}
A transverse magnetization is introduced into the system that oscillates at radio frequency (RF) in order to modify the net magnetization, this is called the RF pulse. When exposing the nuclear spin system to energy of the RF pulse, the equilibrium of the spin system is disturbed. The RF pulse is designed to introduce energy E such that the net spin direction orients at $90^\circ$. The net magnetization now is denoted as $M_{xy}$ as it aligns with the xy-plane instead of the z-axis. Individual spins now precess in phase with the RF pulse, it gives rises to the MR signal in the receiver coil.

@@

\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=\textwidth]{mri/rf-pulse.pdf}
\end{tabular}
\caption{Illustration of the process of transverse relaxation. Initially, the spin precesses about the B0 field under steady-state. When a 90deg RF pulse is applied, the spin orients towards the xy-plane. When the RF pulse is switched off, the spin now relaxes towards the z-axis.  } \label{fig:energy}
\end{center}
\end{figure}

@@

T2 relaxation is so named after the time constant of phase decoherance of the transverse magnetic moment.

@@

Phase coherence is lost in two ways:
\begin{enumerate}
\item Spins are associated with local magnetic fields, and interact randomly The energy transfer that occur in the local magnetic field interactions cause a mumulative loss of phase. It's a pure spin-spin interaction, and thereby not affected by $180^\circ$ refocusing pulse. It is also independent from the strength of $B_0$. 
The time constant which describes the return to equilibrium of the transverse magnetization, is called the \textit{spin-spin
relaxation time} or T2, and is defined in respect to time $t$ as:
$$M_{xy} =M_{xymax} e^{-t/T2}$$
where T2 is the time it takes for the transverse magnetization to decay to 37$\%$ of its original magnitude. 
T2 is always less than or equal to T1.

@@

\begin{figure}[h]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=5.5in]{mri/T1T2}
\end{tabular}
\caption{Diagram depicting the relaxation processes of T1 (panel A), T2 (panel B) and $T_2^*$ (panel C). $T_2^*$ is shorter than T2.} \label{fig:T1T2}
\end{center}
\end{figure}

@@

\paragraph{Repetition Time}
T1 contrast is adjusted by tuning repetition time (TR), which is the length of relaxation interval between excitation pulses. When TR is long (> 1500 ms), $M_z$ recovers totally, and therefore subsequent excitation will again retain maximal signal strength; when the TR is short (< 600 ms), the $M_z$ has not totally recovered, and the next excitation pulse will not be able to gain strong MR signal. Different tissues have different intrinsic T1 differences, and therefore will show strong contrast under short TR. Therefore short TR are said to be T1 weighted.

\paragraph{Echo Time}
MR sequences have a gradient field which is used to control field  inhomogenieties and to encode spatial origin. The gradients also contribute to spin dephasing, therefore a refocusing pulse is necessary to adequately measure MR signals. The time delay between refocusing, and the measured signal under phase coherence (spin echo) is known as echo time (TE). When TE is short ($<$ 30 ms), there is little signal differences between tissue types; when TE is long ( $\geq$ 60 ms ), the intrinsic tissue differences will allow good contrast in the resulting image. Therefore long TE is said to be T2 weighted.

\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=\textwidth]{mri/spin-echo.pdf}
\end{tabular}
\caption{Illustration of the process of spin-echo. Initially, the spin precesses about the B0 field under steady-state. When a 90deg RF pulse is applied, the spins orients towards the xy-plane. When the RF pulse is switched off, the spins now relaxes towards the z-axis, and the spin undergoes phase decoherance. A 180 degrees flip angle pulse is applied to prevent spin saturation, and then a refocusing pulse is applied to bring spins back into coherance.  } \label{fig:spin-echo}
\end{center}
\end{figure}

Typically, T1-weighted spin echo (SE)  sequence is acquired with TR/TE at 340/13 ms; T2-weighted fast spin echo (FSE) at TR/TE of 3500/120 ms. 
In the case of T1-weighted MRI, images are created with short TE and short TR times. They can differentiate fat from water - with fat brighter and water darker (\url{www.mr-tip.com}) by using a gradient echo (GRE) sequence. In the brain T1-weighted scans provide good gray matter/white matter contrast. In addition, the image contrast can be increased with the use of an inversion pulse as in an MP-RAGE sequence. Due to short TR, this scan can be run very quickly, allowing the collection of high resolution 3D structural image (Figure \ref{fig:T1T2PD}, Left). These structural images can provide detailed information of the brain's shape and size, and provide the base for cortical surface reconstruction and segmentation.

@@

\paragraph{Slice Selection} Recall that the resonance condition can only be satisfied when the RF pulse is equal to the Larmor frequency of the nucleus precession under the magnetic field. 
Therefore, to allow slice discrimination, the magnetic field along the z-axis is made to be inhomogeneous by introducing a linear gradient field. It's an extra coil that adds or subtracts from the main magnetic field. With the linear gradient, only a particular portion of the z-axis would share a similar unique Larmor frequency, and can be selectively excited with an RF pulse. The steepness of the gradient determines the slice thickness, the steeper the gradient, the thinner the slices.

@@

\begin{itemize}
\item Phase encoding: The phase encoding gradient is applied in the y-direction. It alters the Larmor frequency of the spins based the location of the y-axis, and results in phase shifts of the spins relative to each other. The phase shift is determined by the duration and amplitude of the gradient, when the gradient is switched off, all spins return to their initial rate of precession, but maintains the relative phase shift. 
\item Frequency encoding: The frequency encoding gradient is applied on the x-direction. This applies a magnetic field that increase in strength from right to left. The change in Larmor frequency alters the precession speed of the spins such that the spins towards the right precess faster than that of the left. The MR signal collected represents the entire frequency spectrum of the spins. 
\end{itemize}

The measured MR signal with both encoding methods allows recovery of spatial information from the scanned area. Application of Fourier transform will decompose the signal into component frequencies along the frequency encoding direction (x). The phase distribution along each frequencies provide information on the phase encoding direction (y). Decomposing the phase distribution requires another Fourier transform on the multiple scans of different phase gradient strengths. This is called two-dimensional Fourier transform (2D-FT), the number of phase gradient measures is called the phase encoding step.

@@

The collected MR data are encoded in a mathematical space termed the k-space. The k-space is a two dimensional feature space, where the x-axis $k_x$ represent the measured frequency domain, and $k_y$ the phase domain. Each step in $k_y$ represents a single phase encoding step. The x-axis origin $k_x=0$ is the gradient iso-center, and the frequency distribution is symmetric about the x-origin. The 2D-FT operation is applied to the k-space to transform the frequency domain information into image.

>>>>

%\graphicspath{{D:/project/thesis_preparation/PhDtemplateLATEX_V2/ch_Background_Work/figures/}}
\section{Magnetic Resonance Imaging}
Magnetic Resonance Imaging (MRI) is an imaging technology that allows non-invasive visualization of the body using magnetic fields. It is a crucial enabler of modern healthcare and medical research and is used for the detection and diagnosis of neuropathology.

MRI is based on the principles of nuclear magnetic resonance (NMR), where atomic nuclei, when placed in a magnetic field, can absorb and emit electromagnetic radiation. An MRI machine applies a primary magnetic field (B0) to magnetize and align the hydrogen nuclei in the body. A radio frequency (RF) field is subsequently introduced perpendicular to B0, and reorients the spin axis. This causes the nuclei to produce NMR signals detectable by the MRI scanner. Based on the precession frequency decay, the signals are recorded to construct an image of the scanned area based on the contrasts in water and fat concentrations.

@@

\begin{figure}[ht]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=3.5in]{mri/spin.pdf}
\end{tabular}
\caption{Spin of a nucleus has a magnetic moment vector $\mu$, which behaves similar to a magnet with north (N) and south (S) poles.}
\label{fig:spin}
\end{center}
\end{figure}

@@

There are more protons with spins with parallel orientation than anti-parallel, due to the increased energy stability. This number of difference is called spin excess and the magnetic moments of the excess spins add to form the net magnetization vector $M_z$. When the spin system is at equilibrium, net magnetization lies along the z axis, parallel to $B_0$.

@@

\subsubsection{Relaxation}
A transverse magnetization is introduced into the system that oscillates at radio frequency (RF) in order to modify the net magnetization, this is called the RF pulse. When exposing the nuclear spin system to the energy of the RF pulse, the equilibrium of the spin system is disturbed. The RF pulse is designed to introduce energy E such that the net spin direction orients at $90^\circ$. The net magnetization now is denoted as $M_{xy}$ as it aligns with the xy-plane instead of the z-axis. Individual spins now precess in phase with the RF pulse, it gives rises to the MR signal in the receiver coil.

@@

\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=\textwidth]{mri/rf-pulse.pdf}
\end{tabular}
\caption{Illustration of the process of transverse relaxation. Initially, the spin precesses about the B0 field under steady-state. When a 90deg RF pulse is applied, the spin orients towards the xy-plane. When the RF pulse is switched off, the spin now relaxes towards the z-axis.  } \label{fig:relaxation}
\end{center}
\end{figure}

@@

T2 relaxation is so named after the time constant of phase decoherence of the transverse magnetic moment.

@@

Phase coherence is lost in two ways:
\begin{enumerate}
\item Spins are associated with local magnetic fields, and interact randomly. The energy transfers that occur in the local magnetic field interactions cause a cumulative loss of phase. These are pure spin-spin interactions, and thereby not affected by $180^\circ$ refocusing pulse. They are also independent of the strength of $B_0$. 
The time constant which describes the return to equilibrium of the transverse magnetization is called the \textit{spin-spin
relaxation time} or T2, and is defined in respect to time $t$ as:
$$M_{xy} =M_{xymax} e^{-t/T2}$$
where T2 is the duration of time for the transverse magnetization to decay to 37$\%$ of its original magnitude. 
T2 is always less than or equal to T1.

@@

\begin{figure}[ht]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=5.5in]{mri/T1T2}
\end{tabular}
\caption{Diagram depicting the relaxation processes of T1 (panel A), T2 (panel B) and $T_2^*$ (panel C). $T_2^*$ is shorter than T2.} \label{fig:T1T2}
\end{center}
\end{figure}

@@

\paragraph{Repetition Time}
T1 contrast is adjusted by tuning repetition time (TR), which is the length of relaxation interval between excitation pulses. When TR is long (> 1500 ms), $M_z$ recovers totally, and therefore subsequent excitation will again retain maximal signal strength. When the TR is short (< 600 ms), the $M_z$ has not entirely recovered, and the next excitation pulse will not be able to regain signal strength. Different tissues have different intrinsic T1 differences, and will show strong contrast under short TR. Therefore short TR are said to be T1 weighted.

\paragraph{Echo Time}
MR sequences have a gradient field which is used to control field inhomogeneities and to encode spatial origin. The gradients also contribute to spin dephasing. Therefore a refocusing pulse is necessary to measure MR signals adequately. The time delay between refocusing and the measured signal under phase coherence (spin echo) is known as echo time (TE). When TE is short ($<$ 30 ms), there are little signal differences between tissue types; when TE is long ( $\geq$ 60 ms ), the intrinsic tissue differences will allow good contrast in the resulting image. Therefore long TE is said to be T2 weighted.

\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
\includegraphics[width=\textwidth]{mri/spin-echo.pdf}
\end{tabular}
\caption{Illustration of the process of spin-echo. Initially, the spin precesses about the B0 field under steady-state. When a 90deg RF pulse is applied, the spins orients towards the xy-plane. When the RF pulse is switched off, the spins now relaxes towards the z-axis, and the spin undergoes phase decoherence. A 180 degrees flip angle pulse is applied to prevent spin saturation, and then a refocusing pulse is applied to bring spins back into coherence.  } \label{fig:spin-echo}
\end{center}
\end{figure}

Typically, T1-weighted spin echo (SE)  sequence is acquired with TR/TE at 340/13 ms; T2-weighted fast spin echo (FSE) at TR/TE of 3500/120 ms. 
In the case of T1-weighted MRI, images are created with short TE and short TR times. They can differentiate fat from water - with fat brighter and water darker (\url{www.mr-tip.com}) by using a gradient echo (GRE) sequence. In the brain T1-weighted scans provide good gray matter/white matter contrast. In addition, the image contrast can be increased with the use of an inversion pulse as in an MP-RAGE sequence. Due to short TR, this scan can be run very quickly, allowing the collection of high resolution 3D structural image (Figure \ref{fig:T1T2PD}, Left). These structural images can provide detailed information on the brain's shape and size, and provide the base for cortical surface reconstruction and segmentation.

@@

\paragraph{Slice Selection} Recall that the resonance condition can only be satisfied when the RF pulse is equal to the Larmor frequency of the nucleus precession under the magnetic field. 
Therefore, to allow slice discrimination, the magnetic field along the z-axis is made to be inhomogeneous by introducing a linear gradient field. It is an extra coil that adds or subtracts from the main magnetic field. With the linear gradient, only a particular portion of the z-axis would share a similar unique Larmor frequency, and can be selectively excited with an RF pulse. The steepness of the gradient determines the slice thickness, the steeper the gradient, the thinner the slices.

@@

\begin{itemize}
\item Phase encoding: The phase encoding gradient is applied in the y-direction. It alters the Larmor frequency of the spins based the location of the y-axis, and results in phase shifts of the spins relative to each other. The phase shift is determined by the duration and amplitude of the gradient, when the gradient is switched off, all spins return to their initial rate of precession, but maintains the relative phase shift. 
\item Frequency encoding: The frequency encoding gradient is applied on the x-direction. This applies a magnetic field that increases in strength from right to left. The change in Larmor frequency alters the precession speed of the spins such that the spins towards the right precess faster than that of the left. The MR signal collected represents the entire frequency spectrum of the spins. 
\end{itemize}

The measured MR signal with both encoding methods allows recovery of spatial information from the scanned area. Application of Fourier transform will decompose the signal into component frequencies along the frequency encoding direction (x). The phase distribution along each frequency provide information on the phase encoding direction (y). Decomposing the phase distribution requires another Fourier transform on the multiple scans of different phase gradient strengths. This is called two-dimensional Fourier transform (2D-FT), the number of phase gradient measures is called the phase encoding step.

@@

The collected MR data are encoded in a mathematical space termed the k-space. The k-space is a two-dimensional feature space, where the x-axis $k_x$ represent the measured frequency domain, and $k_y$ the phase domain. Each step in $k_y$ represents a single phase encoding step. The x-axis origin $k_x=0$ is the gradient iso-center, and the frequency distribution is symmetric about the x-origin. The 2D-FT operation is applied to the k-space to transform the frequency domain information into an image.

### 2017-12-02 thesis/contents/reviews/multiplesclerosis.tex
checkpoint on edit

The MS lesions or plaques are characterized as focal hyper intensities in T2-weighted or proton-density images. Chronic and acute lesions could be distinguished by the presence of blood-brain-barrier (BBB) breakdown with the use of MR contrast agents such as gadolinium. Contrast agent usage however needs to be carefully tuned with the image acquisition parameters.
Focal MS lesions commonly appear as round hyperintensities in MR images ranging from a few milimeters to centimeter in size. They are also found densely dispersed in paraventricular white matter. The presence of MR hyperintensities is not unique to MS, therefore the location, shape and time dissemination must be considered when making a diagnosis. The MAGNIMS guidelines for space dissemination recommends the identification of T2 lesions in at least two of five locations (cortical/juxtacortical, periventricular, infratentorial, optic nerve, and spinal cord) \cite{Filippi2016}. Paraventricular region require at least three or more lesions to be confirmed. Time dissemination recommends the presence of at least two new lesions in a follow-up MRI when compared to a baseline, or the simultaneous presence of asymptomatic lesions.

@@

\subsection{Pain and MS}
Pain is a common co-morbidity of MS, with up to 50-85\% prevalence \cite{Osterberg2005}. Common pain symptoms include optic neuritis, painful tonic seizure and trigeminal neuralgia. 91\% of the patients report pain at time of evaluation, and most experience daily pain. Both central and peripheral pain are reports in people of MS. Central pain in MS is reported to occur 27-58\% of the subjects \cite{O'Connor2008}. The most common central pain in MS are extremity pain, TN, and Lhermitte’s sign. TN has a prevalence 2\% in the MS, which is 20 times higher than in a general population.

TN secondary to MS, or symptomatic TN (MS-TN), is indistinguishable from TN in terms of clinical presentation. It does however present with a demonstrable structural lesion other than vascular compression of CN V. MS-TN has a prevalence rate of about 2-4\% and in some studies, there are reports of prevalence rate as high as 6\%. Unilateral MS-TN is the most common pain type, however there is up to 30\% occurrences of bilateral pain.

>>>>

The MS lesions or plaques are characterized as focal hyperintensities in T2-weighted or proton-density images. Chronic and acute lesions could be distinguished by the presence of blood-brain-barrier (BBB) breakdown with the use of MR contrast agents such as gadolinium. Focal MS lesions commonly appear as round hyperintensities in MR images ranging from a few milimeters to centimeter in size. They are also found densely dispersed in paraventricular white matter. The presence of MR hyperintensities is not unique to MS, therefore the location, shape and time dissemination must be considered when making a diagnosis. The MAGNIMS guidelines for space dissemination recommends the identification of T2 lesions in at least two of five locations (cortical/juxtacortical, periventricular, infratentorial, optic nerve, and spinal cord) \cite{Filippi2016}. Paraventricular region requires at least three or more lesions to be confirmed. Time dissemination recommends the presence of at least two new lesions in a follow-up MRI when compared to a baseline, or the simultaneous presence of asymptomatic lesions.

@@

\subsection{Pain and MS}
Pain is a common co-morbidity of MS, with up to 50-85\% prevalence \cite{Osterberg2005}. Common pain symptoms include acute pain such as optic neuritis and painful tonic seizure, as well as chronic pain such as trigeminal neuralgia. 91\% of the patients report pain at the time of evaluation, and most experience daily pain. Both central and peripheral pain are reports in people of MS. Central pain in MS is reported to occur 27-58\% of the subjects
\cite{OConnor2008}. The most common pain in MS include extremity pain and TN. TN has a prevalence 2\% in the MS, which is 20 times higher than in a general population.

TN secondary to MS, or symptomatic TN (MS-TN), is frequently indistinguishable from TN in terms of clinical presentation. It does however present with a demonstrable structural lesion other than vascular compression of CN V. MS-TN has a prevalence rate of about 2-4\% and in some studies, there are reports of prevalence rate as high as 6\%. Unilateral MS-TN is the most common pain type, however there is up to 30\% occurrences of bilateral pain.

### 2017-12-02 thesis/contents/reviews/pain.tex
checkpoint on edit

Pain is an unique and important sensory dimension that is central to the evolution and survival of animal species. In humans, the sensation of pain is generally described as having an unpleasant quality. Additionally, the experience of pain is complex, consisting of nociceptive, emotional and cognitive processes. The international association for the study of pain (IASP) defines pain as "An unpleasant sensory and emotional experience associated with actual or potential tissue damage, or described in terms of such damage"\footnote{http://www.iasp-pain.org/Taxonomy\#Pain}.

@@

The gate-control theory \cite{Melzack1965a} (\ref{fig:gate-theory}), which forms the basis of modern theory of pain, postulates that the cells of the substantia gelatinosa selectively inhibit primary afferent into the central transmission cells that relays touch or pain sensations. The transmission control depends on the large and small diameter primary afferents, where large fibers inhibit, while small fibers facilitate the transmission. The gate-control theory reconciles the two previously divergent specificity and pattern theories of pain.

@@

The concept of a pain specific pathway was first put forward by Charles Bell where he proposed that the brain is a heterogeneous structure, and that nerves are a mix of heterogeneous neurons that became bundled to facilitate anatomical distribution \cite{Bell1868}. Evidence soon followed with the discovery of specific touch receptors and the demonstration that resection of posterior spinal tracts in animals by Schiff and Woroschiloff isolated innocuous touch from pain and temperature conductance (anterolateral tract) \cite{Dallenbach1939}. Von Frey demonstrated distinct receptive fields that respond to pressure that triggered tactile, warm, cold and pain. Sherrington further developed the specificity theory by coining the term “nocicipient” and unified attributes of intensity theory and specificity theory.

@@

The key contribution of the gate-control theory was to emphasize the role of the dorsal horn in convergent pain processing, and therefore reconciles the evidences of pain-specific primary afferent and with pattern theory. The current theory of pain is defined such that the spatial and temporal profile of peripheral nerve activity encodes all the necessary information for pain. That additional processing of information in the periphearla nd central nervous system, both divergent and convergent, are in the form of patterns of cyclic nervous activities. This entire pattern of neural activities is the pain sensory experience. Specifically, peripheral somatosensory and nociceptive pain afferents transmit to the spinal dorsal horn ( laminae V ), whereby wide-dynamic-range (WDR) neurons convergently process these primary afferent, and relay them to the thalamus, and subsequently the primary sensory cortex. A central concept, which was also introduced by Melzacks, is the pain neuromatrix. Which dictates further cyclically convergent and divergent cortical patterns involving multiple cortical regions that in its entirety defines the entire pain perception \cite{Melzack1999}. (See section \ref{section:neuromatrix})

@@

\paragraph{Default mode network}
The default mode network (DMN) are regions of the brain, including the cingulate cortex, which show highly correlative BOLD activations in fMRI during wakeful rest \cite{Grecius2003,Raichle2015}. The core regions associated with the default network are ventral medial prefrontal cortex (vmPFC), posterior cingulate, inferior parietal, lateral temporal cortex, dorsal medial prefrontal cortex (dMPFC) and hippocampal formation. Regions such as vmPFC, dMPFC and the posterior cingulate overlap with the affective-motivational pain matrix, and are found to be disrupted in DMN in chronic pain \cite{Baliki2008}. DMN activation has been to correlate with antinociception \cite{Kucyi2013} and it may undergo functional reorganization under chronic pain \cite{Baliki2014}.

@@

Peripheral neuropathic pains can arise when there are disruptions in peripheral signal transmission. The disruption can be a result of injury to the first order peripheral afferent neurons, which result in increased sensitivity to stimuli and via ectopic action potentials. Changes in peripheral neurotropic factors as a result of an injury can cause increased membrane excitability or spurious sprouting of the synaptic terminals in the dorsal horn of the spinal cord. Additionally, maladaptive changes may occur in the secondary sensory neurons and interneurons in the dorsal horn laminae that amplifies or reroutes the noxious signals.

@@

Neuropathic pain can also arise from changes in the central neuroaxis. Trauma and other disruptions to the spinal pathways can affect noxious signal transmission. White matter pathways within the brain may also become damaged by conditions such as trauma, infections, and multiple sclerosis, such that the synchronicity of action potential conduction between cortical and subcortical structures are disrupted, and thereby give rise to neuropathic pain. It is believed that injury to white matter fibers can affect action potential conduction between cortical and subcortical structures. This may give rise or contribute to neuropathic pain (ref).

 The utilization of magnetic resonance imaging (MRI), coupled with novel imaging modalities such as functional MRI (fMRI) and diffusion MRI (dMRI), have been instrumental in our understanding of pain, permitting both \textit{in vivo} measures of brain activity during pain (functional studies), as well as assessment of changes to the brain microstructure.

>>>>

Trigeminal Neuralgia is a chronic facial pain syndrome. It is therefore imperative to review the evolving history and theories of pain. Pain is an unique and important sensory dimension that is central to the evolution and survival of animal species. In humans, the sensation of pain is generally described as having an unpleasant quality. Additionally, the experience of pain is complex, consisting of nociceptive, emotional and cognitive processes. The international association for the study of pain (IASP) defines pain as "An unpleasant sensory and emotional experience associated with actual or potential tissue damage, or described in terms of such damage"\footnote{http://www.iasp-pain.org/Taxonomy\#Pain}.

@@

The gate-control theory \cite{Melzack1965a} (\ref{fig:gate-theory}), which forms the basis of modern theory of pain, postulates that the cells of the substantia gelatinosa selectively inhibit signal transmission from primary afferent to the central transmission cells that relays touch or pain sensations. The transmission control depends on the large and small diameter primary afferents, where large fibers inhibit, while small fibers facilitate the transmission. The gate-control theory reconciles the two previously divergent specificity and pattern theories of pain.

@@

The concept of a pain specific pathway was first put forward by Charles Bell where he proposed that the brain is a heterogeneous structure, and that nerves are a mix of heterogeneous neurons that became bundled to facilitate anatomical distribution \cite{Bell1868}. Evidence soon followed with the discovery of specific touch receptors and the demonstration that resection of posterior spinal tracts in animals by Schiff and Woroschiloff isolated innocuous touch from pain and temperature conductance (anterolateral tract) \cite{Dallenbach1939}. Von Frey demonstrated distinct receptive fields that respond to pressure that triggered tactile, warm, cold and pain. Sherrington further developed the specificity theory by coining the term “nocicipient” and unified attributes of intensity theory and specificity theory \cite{Casey1982}.

@@

The key contribution of the gate-control theory was to emphasize the role of the dorsal horn in convergent pain processing, and therefore reconciles the evidences of pain-specific primary afferent and with pattern theory. The current theory of pain is defined such that the spatial and temporal profile of peripheral nerve activity encodes all the necessary information for pain. That additional processing of information in the peripheral and central nervous system, both divergent and convergent, are in the form of patterns of cyclic nervous activities. This entire pattern of neural activities is the  sensory pain experience. Specifically, peripheral somatosensory and nociceptive pain afferents transmit to the spinal dorsal horn ( laminae V ), whereby wide-dynamic-range (WDR) neurons convergently process these primary afferents, and relay them to the thalamus, and subsequently the primary sensory cortex. A central concept, which was also introduced by Melzacks, is the pain neuromatrix, which dictates further cyclically convergent and divergent cortical patterns involving multiple cortical regions that in its entirety defines the entire pain perception \cite{Melzack1999}. (See section \ref{section:neuromatrix})

@@

\paragraph{Default mode network}
The default mode network (DMN) are regions of the brain, including the cingulate cortex, which show highly correlative BOLD activations in fMRI during wakeful rest \cite{Greicius2003,Raichle2015}. The core regions associated with the default network are ventral medial prefrontal cortex (vmPFC), posterior cingulate, inferior parietal, lateral temporal cortex, dorsal medial prefrontal cortex (dMPFC) and hippocampal formation. Regions such as vmPFC, dMPFC and the posterior cingulate overlap with the affective-motivational pain matrix, and are found to be disrupted in DMN in chronic pain \cite{Baliki2008}. DMN activation has been to correlate with antinociception \cite{Kucyi2013} and it may undergo functional reorganization under chronic pain \cite{Baliki2014}.

@@

Peripheral neuropathic pains can arise when there are disruptions in peripheral signal transmission. The disruption can be a result of injury to the first order peripheral afferent neurons, which result in increased sensitivity to stimuli and via ectopic action potentials. Changes in peripheral neurotropic factors as a result of an injury can cause increased membrane excitability or spurious sprouting of the synaptic terminals in the dorsal horn of the spinal cord. Additionally, maladaptive changes may occur in the secondary sensory neurons and interneurons in the dorsal-horn laminae that amplifies or reroutes the noxious signals \cite{Elmes2004}.

@@

Neuropathic pain can also arise from changes in the central neuroaxis. Trauma and other disruptions to the spinal pathways can affect noxious signal transmission. White matter pathways within the brain may also become damaged by conditions such as trauma, infections, and multiple sclerosis, such that the synchronicity of action potential conduction between cortical and subcortical structures are disrupted, and thereby give rise to neuropathic pain. It is believed that injury to white matter fibers can affect action potential conduction between cortical and subcortical structures. This may give rise or contribute to neuropathic pain \cite{Geha2008}.

 The utilization of magnetic resonance imaging (MRI), coupled with novel imaging modalities such as functional MRI (fMRI) and diffusion MRI (dMRI), have been instrumental in our understanding of pain, permitting both \textit{in vivo} measures of brain activity during pain (functional studies) \cite{Davis2006}, as well as assessment of changes to the brain microstructure \cite{Hodaie2009a,Chen2015c}.
